---
title: "Exploring"
author: "Camila Vargas"
date: "8/3/2018"
output: html_document
---
Exploratory data analysis, or EDA - explore your data in a systematic way

Cycle:
1. Generate questions about your data.

2. Search for answers by visualising, transforming, and modelling your data.

3. Use what you learn to refine your questions and/or generate new questions.

Data cleaning:
Ask questions about whether your data meets your expectations or not. To do data cleaning, you’ll need to deploy all the tools of EDA: visualisation, transformation, and modelling.

Package:
```{r}
library(tidyverse)
```

Goal during EDA is to develop an understanding of your data!
Its a creative process. The important thing is to ask right qustions.

Two leading questions:
1. What type of variation occurs within my variables?

2. What type of covariation occurs between my variables?


Distribution of categorical variables using a bar graph

```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```

To count the number of count in each bar

```{r}
diamonds %>% 
  count(cut)
```

To examine the distribution of continous variables use a histogram

```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)
```

To compute this counts you combine: ´dplyr::count()´ and ´ggplot2::cut_width()´


```{r}
diamonds %>% #from the data base dimonds
  count(cut_width(carat, 0.5)) #