---
title: "ggplot practice"
author: "Camila Vargas"
date: "5/3/2018"
output: html_document
---

ggplot practcice

```{r}
library(tidyverse)
ggplot(data = mpg, aes(x=displ, y=hwy))+
  geom_point()
```
 
Or, you can also say to ggplot to map witing the geom_point function

```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x=displ, y=hwy))
```


#Exercise 3.2.4

1. Run ggplot(data = mpg). What do you see?
```{r}
ggplot(data = mpg)
```

By plottong just ggplot with the data, R gives you juts the layout for the graph

2. How many rows are in mpg? How many columns?

Using the str() function, str(mpg) you can explore the variables and observations of an object.
Inn this case there are 234 rows and 11 columnes

3. What does the drv variable describe? Read the help for ?mpg to find out.
The drv variable describes the type of drive the car has. The possible categories in this variables are: f = front-wheel drive, r = rear wheel drive, 4 = 4wd

4. Make a scatterplot of hwy vs cyl.
```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x=hwy, y=cyl))
```

5. What happens if you make a scatterplot of class vs drv? Why is the plot not useful?

```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x=class, y=drv))
```

This is not a useful graph because class and drv are categorical variables. 

********

An aesthetic is a visual property of the objects in your plot. Aesthetics include things like the size, the shape, or the color of your points.


QUESTION: How can you color just one category of your variable???
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = compact))
```

To set an aesthetic manually, set the aesthetic by name as an argument of your geom function; i.e. it goes outside of aes(). You’ll need to pick a value that makes sense for that aesthetic:

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), shape = 22)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), shape = 22, color="red")

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), shape = 22, fill="red")
```

 When the color, shape, fill functions are outside the aes then it changes the appereance of all the grpah.
 
 
#Exercise 3.3.1

2. Which variables in mpg are categorical? Which variables are continuous? (Hint: type ?mpg to read the documentation for the dataset). How can you see this information when you run mpg?

This data set has 11 variables:
Manufacture moder: categorical
displ, engine displacement: continuous
year: ?? categorical
cyl, number of cylinders: ??continuous (class = interger)
trans: categorical
dvr: categorical
cty: continuous
hwy: continuous
fl, fuel type: categorical
class, type of car: categorical

QUESRION: How can you see this information when you run mpg?????


3. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?


4. What happens if you map the same variable to multiple aesthetics?

5. What does the stroke aesthetic do? What shapes does it work with? (Hint: use ?geom_point)
 
 
 
 
 



